# Annualizing benefits and costs

**This module will teach you how to use the `ecorest` and `EngrEcon` packages in `R` to efficiently and reproducibly calculated annualized benefits and costs for restoration project planning:**

Annualization is a technique for determining the costs and benefits of projects on an annual time scale, regardless of the overall time horizon of the project. Calculating annualized costs and benefits is an important component of restoration planning because it enables a fair comparison between alternatives project whose costs and/or benefits occur at different periods of time following implementtion. 

## Annualizing benefits
Annualization of benefits can be easily carried out in the `ecorest` package. ALong with `ecorest` we will also load the `tidyverse` package for plotting/wrangling and the `EngrEcon` package for cost annualization.

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
library(ecorest)
library(EngrEcon)
```


Imagine a simple scenario with ecosystem or restoration benefits calculated at four time periods. We can store these benefits in a dataframe and plot them with `ggplot` as follows:

```{r}
#User-specified time intervals
ben_df <- data.frame(year = c(0,2,20,50),
                     ben = c(0,100,90,80))

p <- ggplot(ben_df)+
    geom_point(aes(year, ben), size = 2)+
    geom_line(aes(year, ben))+
  labs(x = "Year", y = "Benefits")+
  theme_classic()
  
p
  
```
What annualization does is calculate the total area under this benefits curve, and then divide by the number of years. This essentially looks like carving the area into triangles and rectangles, calculating the areas of these shapes, and dividing to get an annual benefit.

```{r}
df_rect <- tibble(xmin = ben_df$year[2:3],
                  xmax = ben_df$year[3:4],
                  ymin = c(0,0),
                  ymax = ben_df$ben[3:4])

p + geom_rect(data = df_rect,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = NA, lty = "dashed", color = "black")

```

The `ecorest` package provides an easy way to achieve this with the annualizer function, which computes "time-averaged quantities based on linear interpolation," which does exactly this.

Inputs for the function are: 1) a vector of time intervals and 
2) a vector of values to interpolate>

```{r}
annualizer(ben_df$year, ben_df$ben)
```
This indicates that the annualized ecological benefits of the project are 87.2 units (e.g., acres, habitat units, etc.)

## Annualizing costs

Cost annualization is carried out so that fair comparisons can be made between projects with different temporal scales.

EDIT THE FOLLOWING.

Cost estimates were compiled for each site-scale restoration action following standard cost engineering methods. At this phase, construction cost represents a parametric cost for comparative purposes only, which is confined to construction activities. No real estate, pre-construction engineering and design, construction management, or cultural resources costs were included at present.

Monitoring and adaptive management are typically conducted over a ten-year window for ecosystem restoration projects (WRDA 2007, Section 2036). Monitoring cost was estimated for four points in time, a sampling event during pre-construction engineering design as well as sampling in years 1, 3, and 5 following construction. Each sampling event assumed collection of data on fish communities, invertebrate communities, bathymetric mapping, and hydraulic outcomes. Adaptive management was estimated a proportion of construction cost ranging between 7.5% and 25% depending on site- and alternative-complexity (Appendix B).

Operations, maintenance, repair, replacement, and rehabilitation (OMRRR) costs were estimated relative to four common practices. First, quarterly site visits were planned for general maintenance to any small-scale restoration features, trash removal, and minor maintenance of educational features (signage, benches, or classrooms) at a cost of \$10,000 per year. Second, bi-annual site visits were planned to include a simple inspection, which were estimated as \$4.00 per linear foot of stream every other year. Third, invasive species removal was estimated to cost \$15,000 per site and occur once every 10-years. Fourth, repair of structural restoration features (e.g., rock, wood, and earth work) was assumed to occur once every 25 years. These estimates resulted in OMRRR costs ranging from 1.2% to 16.7% or construction cost depending on the site and alternative (Appendix B, Table B1), which is approximately in line with other stream restoration projects (Abera and McKay 2023).

Average annual economic costs were computed for all cost categories. Interest during construction was computed based on construction costs with site- and alternative-specific construction durations. The FY24 Federal discount rate (2.75%, USACE 2023) was used to annualize construction cost, interest during construction, and monitoring and adaptive management expenses over a 50-year planning horizon. Cost data were annualized using the [EngrEcon](https://cran.r-project.org/web/packages/EngrEcon/index.html) software package, although a [web application](https://wrises.shinyapps.io/engrecon-webapp/) also exists for conducting these calculations. Table 3 provides an example of cost estimates for Site-17F2M, and Appendix B provides a cost summary for all sites and actions.

Cost data were annualized using the EngrEcon software package, although a web application also exists for conducting these calculations.

### Steps for annualization:

Read in initial costs: Project_first, Monitoring, AdaptiveMan

```{r, message = F}
#Populate these sections of code from code below

# Set federal discont rate for water resources projects in FY24
discount <- 0.0275

# Read in initial costs
costs <- read_csv("data/utoy_cost.csv")

```

Calculate present value of OMRR costs using `om_distribute` function. This function...
```{r}

costs_omrrr <- costs %>%
  rowwise() %>%
  mutate(General = om_distribute(discount, 50, 1, 10000 * ifelse(RipAreaFt2>0, 1, 0)),
         InspectionTrash = om_distribute(discount, 50, 2, 4*ReachLengthFt),
         Invasives = om_distribute(discount, 50, 10, 15000 * ifelse(RipAreaFt2>0, 1, 0)),
         WoodRockEarth = om_distribute(discount, 50, 25, 0.1*ProjectFirst),
         TotalOMRR = InspectionTrash + General + Invasives + WoodRockEarth)

```

Now we have the present value of all costs in one dataframe: Project_first, Monitoring, AdapMan the same; OMRR costs from previous step


Calculate Interest during construction using  based on discount rate, duration of construction, and construction cost using the interest_during_construction function
```{r}
costs_idc <- costs_omrrr %>%
  rowwise() %>%
  mutate(IDC = case_when(DurationMon==0~0, TRUE ~ interest_during_construction(discount, DurationMon, ProjectFirst)))

```

Last step: annualize all costs using present_to_annual function using discount rate and 50 year time horizon
```{r}
costs_ann <- costs_idc %>% 
  rowwise() %>%
  mutate(ProjectFirstAnn = present_to_annual(discount, 50, ProjectFirst),
         IDCAnn = present_to_annual(discount, 50, IDC),
         MonitoringAnn = present_to_annual(discount, 50, Monitoring),
         AdManAnn = present_to_annual(discount, 50, AdMan),
         OMAnn = present_to_annual(discount, 50, TotalOMRR),
         TotalAnn = ProjectFirstAnn + MonitoringAnn + AdManAnn + IDCAnn + OMAnn)

```

Finally, we will create a table to display a subset of the results.
```{r}
Table1 <- cbind(costs_ann$ReachID, costs_ann$SiteAction, costs_ann$DurationMon,
                 formatC(costs_ann$ProjectFirst, format="f", digits=0, big.mark = ","),
                 formatC(costs_ann$Monitoring, format="f", digits=0, big.mark = ","),
                 formatC(costs_ann$AdMan, format="f", digits=0, big.mark = ","),
                 formatC(costs_ann$OMAnn, format="f", digits=0, big.mark = ","),
                 formatC(costs_ann$TotalAnn, format="f", digits=0, big.mark = ","))
Table1 <- Table1[1:4,]

colnames(Table1) <- c("Site", "Alternative", "Construction Duration (mo)", 
                       "Construction Cost", "Monitoring Cost", "Adapative Management Cost",
                       "OMRRR Annual Cost", "Total Annualized Cost")

knitr::kable(Table1, caption="Table 1. Example of monetary cost data for Site-17F2M.", align='c') 

```
